---
title: "HCV"
author: "Mathieu"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output:
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
library(data.table)
library(ggplot2)
library(tidyr)
library(dplyr)
library(readr)
library(knitr)
library(zeallot)
library(stringr)
library(lemon)
library(skyplotr)

knitr::opts_chunk$set(echo = FALSE, message = FALSE, fig.path = "Figs/HCV/")
```

```{r}
source("R/utils.R")
source("R/renamer.R")
```

\newpage


```{r}
parse_beast <- function(log_file, model, burnin = 0.1) {
  df <- fread(log_file,
    sep = "\t",
    skip = "#",
    showProgress = FALSE
  )

  state_count <- df$state[nrow(df)]
  df <- myrename(df, df.rename.beast) %>% slice(as.integer(burnin * nrow(df)):n())

  beast_logtheta <- df %>% select(starts_with("skygrid.logPopSize"))
  sky <- prepare_skygrid(df$rootHeight,
    beast_logtheta,
    400,
    age_of_youngest = 1994
  )
  sky <- sky %>% mutate(
    divergence = NA,
    algorithm = "MCMC",
    model = model,
    program = "BEAST",
    engine = "BEAGLE",
    label = paste0("BEAST-", model),
    label2 = "BEAST MCMC"
  )

  row_sum <- as.numeric(rowSums(select(df, AC, AG, AT, CG, CT, GT)))
  df_params <- df %>%
    mutate(across(c("AC", "AG", "AT", "CG", "CT", "GT"), ~ . / row_sum)) %>%
    select(rootHeight, precision, freqA, AC, shape) %>%
    gather("param", "value") %>%
    mutate(
      divergence = NA,
      algorithm = "MCMC",
      model = model,
      program = "BEAST",
      engine = "BEAGLE",
      label = paste0("BEAST-", model),
      label2 = "BEAST MCMC"
    )

  df_probs <- df %>%
    select(likelihood, prior, coalescent) %>%
    gather("param", "value") %>%
    mutate(
      divergence = NA,
      algorithm = "MCMC",
      model = model,
      program = "BEAST",
      engine = "BEAGLE",
      label = paste0("BEAST-", model),
      label2 = "BEAST MCMC"
    )

  time <- parse_time(file.path(dirname(log_file), "beast.log"))

  time <- data.frame(
    time = time,
    model = model,
    iterations = state_count
  )

  list(
    params = df_params,
    probs = df_probs,
    sky = sky,
    time = time
  )
}

human_readable_labels <- expression(
  AC = "A" %<->% "C relative rate",
  rootHeight = "root height",
  precision = "GMRF precision",
  shape = "rate variation shape",
  freqA = pi["A"]
)

human_readable_labeller <- function(x) {
  list(as.list(human_readable_labels)[x$param])
}
```

```{r}
time <- parse_time(file.path("results", "HCV", "beast", "skyglide", "beast.log"))

df <- fread(
  file.path("results", "HCV", "beast", "skyglide", "HCV.log"),
  sep = "\t",
  skip = "#",
  data.table = FALSE,
  showProgress = FALSE
)
lookup <- sapply(1:75, function(x) {
  paste0("skygrid.logPopSize", x)
})
names(lookup) <- sapply(1:75, function(x) {
  paste0("popSize", x)
})
params_id <- c(
  "rootHeight",
  "freqA",
  "AC",
  "shape",
  "precision"
)
df <- myrename(df, df.rename.beast) %>%
  rename(all_of(lookup)) %>%
  select(all_of(params_id), contains("popSize"), "state")

vars <- cvs <- means <- df[FALSE, ]
var0 <- summarize_all(df, var)
sd0 <- sqrt(var0)
mean0 <- summarize_all(df, mean)
cv0 <- sd0 / mean0

every <- 5000
replicates <- nrow(df) / every
sampled_every <- df$state[2] - df$state[1]

for (i in 1:replicates) {
  df_i <- df[1:(i * every), ]
  df2_var <- summarize_all(df_i, var)
  df2_mean <- summarize_all(df_i, mean)
  df2_sd <- sqrt(df2_var)
  df2_cv <- df2_sd / df2_mean
  vars <- bind_rows(vars, (df2_var - var0)**2)
  cvs <- bind_rows(cvs, (df2_cv - cv0)**2)
  means <- bind_rows(means, (df2_mean - mean0)**2)
}

states <- seq(1, (replicates - 1) * every, every) * sampled_every
vars$state <- states
cvs$state <- states
means$state <- states

state_count <- df$state[length(df$state)]

vars <- vars %>% mutate(time = time / state_count * state, program = "BEAST")
cvs <- cvs %>% mutate(time = time / state_count * state, program = "BEAST")
means <- means %>% mutate(time = time / state_count * state, program = "BEAST")
```


```{r}
lookup <- sapply(0:74, function(x) {
  paste0("coalescent.theta.", x)
})
names(lookup) <- sapply(1:75, function(x) {
  paste0("popSize", x)
})

for (divergence in c("ELBO", "KLpq-10")) {
  d <- if (divergence == "KLpq-10") "KL(p||q)" else divergence
  p <- file.path("results", "HCV", "torchtree", "GTR", "skyglide", divergence, "physher")

  time <- parse_time(file.path(p, "torchtree.log"))

  df_m <- fread(file.path(p, "means.csv"),
    sep = ",",
    data.table = FALSE
  )
  df_v <- fread(file.path(p, "variances.csv"),
    sep = ",",
    data.table = FALSE
  )

  df_m <- myrename(df_m, df.rename) %>%
    rename(state = sample) %>%
    rename(all_of(lookup)) %>%
    select(all_of(params_id), contains("popSize"), "state")
  df_v <- myrename(df_v, df.rename) %>%
    rename(state = sample) %>%
    rename(all_of(lookup)) %>%
    select(all_of(params_id), contains("popSize"), "state")

  states <- as.numeric(df_m$state)
  time_per_state <- time / states[length(states)]

  df <- sweep(df_m, 2, as.numeric(df_m[nrow(df_m), ]), "-")**2
  means <- df %>%
    mutate(
      state = states,
      program = d,
      time = time_per_state * state
    ) %>%
    bind_rows(means)

  df <- sweep(df_v, 2, as.numeric(df_v[nrow(df_v), ]), "-")**2
  vars <- df %>%
    mutate(
      state = states,
      program = d,
      time = time_per_state * state
    ) %>%
    bind_rows(vars)

  df <- sweep(sqrt(df_v) / df_m, 2, as.numeric(sqrt(df_v[nrow(df_v), ]) /
    df_m[nrow(df_m), ]), "-")**2
  cvs <- df %>%
    mutate(
      state = states,
      program = d,
      time = time_per_state * state
    ) %>%
    bind_rows(cvs)
}
```

# Squared difference of coefficient of variation against time
```{r}
df2 <- cvs %>%
  select(all_of(c(params_id, "state", "program", "time"))) %>%
  # filter(time > 1000) %>%
  mutate(program = factor(program, levels = c("KL(p||q)", "ELBO", "BEAST"))) %>%
  gather("param", "CV", -state, -time, -program)

g <- ggplot(df2, aes(x = time, y = CV, color = program)) +
  my_theme +
  geom_line() +
  facet_wrap(~param, scales = "free", labeller = human_readable_labeller) +
  labs(x = "Time (minutes)", y = bquote((CV[i] - CV[M])^2), color = NULL)

reposition_legend(g, "center", panel = "panel-3-2")
```

\newpage

# Squared difference of coefficient of variation against time (log scale)
```{r}
df2 <- cvs %>%
  select(all_of(c(params_id, "state", "program", "time"))) %>%
  filter(state > 10000) %>%
  mutate(program = factor(program, levels = c("KL(p||q)", "ELBO", "BEAST"))) %>%
  gather("param", "value", -state, -time, -program) %>%
  filter(!is.infinite(log10(value)))

g <- ggplot(df2, aes(x = time, y = value, color = program)) +
  my_theme +
  geom_line() +
  scale_y_continuous(trans = "log10") +
  facet_wrap(~param, scales = "free", labeller = human_readable_labeller) +
  labs(x = "Time (minutes)", y = bquote((CV[i] - CV[M])^2), color = NULL)

reposition_legend(g, "center", panel = "panel-3-2")
```

\newpage

# Squared difference of coefficient of variation against time (log scale, VB only)
```{r}
df2 <- cvs %>%
  select(all_of(c(params_id, "state", "program", "time"))) %>%
  # filter(state > 10000) %>%
  gather("param", "value", -state, -time, -program) %>%
  filter(program != "BEAST") %>%
  filter(!is.infinite(log10(value)))

g <- ggplot(df2, aes(x = time, y = value, color = program)) +
  my_theme +
  geom_line() +
  # geom_vline(xintercept=as.numeric(df.m$state)[1]) +
  scale_y_continuous(trans = "log10") +
  facet_wrap(~param, scales = "free", labeller = human_readable_labeller) +
  labs(x = "Time (minutes)", y = bquote((CV[i] - CV[M])^2), color = NULL)

reposition_legend(g, "center", panel = "panel-3-2")
```

\newpage

# Squared difference of coefficient of variation against time (BEAST only)
```{r}
df2 <- cvs %>%
  select(all_of(c(params_id, "state", "program", "time"))) %>%
  gather("param", "CV", -state, -time, -program) %>%
  filter(program == "BEAST")

g <- ggplot(df2, aes(x = time, y = CV, color = program)) +
  my_theme +
  geom_line() +
  facet_wrap(~param, scales = "free", labeller = human_readable_labeller) +
  labs(x = "Time (minutes)", y = bquote((CV[i] - CV[M])^2), color = NULL)

reposition_legend(g, "center", panel = "panel-3-2")
```

# Smoothing of trajectory using sliding window
```{r}
df2 <- cvs %>%
  select(!contains("popSize")) %>%
  filter(program == "BEAST") %>%
  mutate(across(
    !c(time, program, state),
    ~ stats::filter(.x, filter = rep(1, 50), side = 2) / 50
  ))

df2 <- cvs %>%
  select(!contains("popSize")) %>%
  filter(program %in% c("ELBO", "KL(p||q)")) %>%
  mutate(across(
    !c(time, program, state),
    ~ stats::filter(.x, filter = rep(1, 100), side = 2) / 100
  )) %>%
  bind_rows(df2)


df2 <- df2[rowSums(is.na(df2)) == 0, ] %>% gather("param", "value", -state, -time, -program)

g <- ggplot(df2, aes(x = time, y = value, color = program)) +
  geom_line() +
  facet_wrap(~param, scales = "free", labeller = human_readable_labeller) +
  labs(x = "Time (minutes)", y = bquote((CV[i] - CV[M])^2), color = NULL)

reposition_legend(g, "center", panel = "panel-3-2")
```

\newpage

```{r}
c(params, probs, skys, times) %<-% parse_beast(file.path("results", "HCV", "beast", "skygrid", "HCV.log"), "skygrid", 0.1)
data <- parse_beast(file.path("results", "HCV", "beast", "skyglide", "HCV.log"), "skyglide", 0.1)
skys <- rbind(skys, data$sky)
params <- rbind(params, data$params)
probs <- rbind(probs, data$probs)
times <- rbind(times, data$time)
times <- times %>% mutate(program = "BEAST", divergence = NA, algorithm = "MCMC", engine = "BEAGLE")

for (model in c("skyglide", "skygrid")) {
  for (divergence in c("ELBO", "KLpq-10")) {
    for (engine in c("physher")) {
      path <- file.path("results", "HCV", "torchtree", "GTR", model, divergence, engine)
      df <- read.csv(file.path(path, "samples.csv"), sep = "\t")
      df <- myrename(df, df.rename)

      df_logtheta <- df %>%
        select(starts_with("coalescent.theta")) %>%
        log()

      skygrid <- prepare_skygrid(df$rootHeight,
        df_logtheta,
        400,
        age_of_youngest = 1994
      )

      d <- if (divergence == "KLpq-10") "KL(p||q)" else divergence

      label <- paste("torchtree", model, d, sep = "-")
      label2 <- paste("torchtree", d)

      skys <- skygrid %>%
        mutate(
          label = label,
          label2 = label2,
          divergence = d,
          model = model,
          program = "torchtree",
          engine = engine,
          algorithm = "VB",
        ) %>%
        bind_rows(skys)

      params <- df %>%
        select(rootHeight, freqA, AC, shape, precision) %>%
        gather("param", "value") %>%
        mutate(
          label = label,
          label2 = label2,
          divergence = d,
          program = "torchtree",
          model = model,
          engine = engine,
          algorithm = "VB",
        ) %>%
        bind_rows(params)

      probs <- df %>%
        select(likelihood, coalescent, prior) %>%
        mutate(prior = prior - 6 * log(6)) %>%
        gather("param", "value") %>%
        mutate(
          label = label,
          label2 = label2,
          divergence = d,
          program = "torchtree",
          model = model,
          engine = engine,
          algorithm = "VB",
        ) %>%
        bind_rows(probs)

      df_info <- read.csv(file.path(path, "info.csv"))
      iterations <- df_info$iters[1]
      time <- df_info$time[1]

      times <- data.frame(
        time = time,
        divergence = d,
        model = model,
        algorithm = "VB",
        program = "torchtree",
        engine = engine,
        iterations = iterations
      ) %>%
        bind_rows(times)
    }
  }
}
skys <- skys %>% mutate(across(c("trajectory_low", "trajectory_high", "trajectory"), exp))
```

```{r}
pp <- file.path("results", "HCV", "torchtree", "GTR")

for (model in c("skyglide", "skygrid")) {
  for (algorithm in c("mcmc", "hmc")) {
    for (engine in c("physher")) {
      label <- paste("torchtree", model, toupper(algorithm), sep = "-")
      label2 <- paste("torchtree", toupper(algorithm))

      p <- file.path(pp, model, algorithm, engine, "samples.csv")

      if (!file.exists(p)) next

      df <- read.csv(p, sep = "\t")
  df <- df[(0.2 * nrow(df)):nrow(df), ]
  df <- myrename(df, df.rename)

  df_logtheta <- df %>% select(starts_with("coalescent.theta")) # %>% log()
  skygrid <- prepare_skygrid(df$rootHeight,
    df_logtheta,
    400,
    age_of_youngest = 1994
  )
  skys <- skygrid %>%
    mutate(
      label = label,
      label2 = label2,
      divergence = NA,
      model = model,
      program = "torchtree",
          engine = engine,
          algorithm = algorithm
    ) %>%
    mutate(across(c("trajectory_low", "trajectory_high", "trajectory"), exp)) %>%
    bind_rows(skys)

  probs_id <- c("likelihood", "prior", "coalescent")

  df_hmc <- df %>%
    select(all_of(
      c(params_id, probs_id)
    )) %>%
    mutate(prior = prior - 6 * log(6)) %>%
    gather("param", "value") %>%
    mutate(
      label = label,
      label2 = label2,
      divergence = NA,
      program = "torchtree",
          engine = engine,
          model = model,
          algorithm = algorithm
    )

  params <- df_hmc %>%
    filter(param %in% params_id) %>%
    bind_rows(params)

  probs <- df_hmc %>%
    filter(param %in% probs_id) %>%
    bind_rows(probs)
    }
  }
}
```

# Skygrid

## Parameter densities with skygrid
```{r skygrid-params}
df_params <- params %>%
  filter(model == "skygrid", engine != "torchtree") %>%
  filter(label2 != "torchtree MCMC") %>%
  mutate(label2 = factor(label2, levels = c("BEAST MCMC", "torchtree HMC", "torchtree ELBO", "torchtree KL(p||q)")))

df_params_beast <- df_params %>% filter(program == "BEAST")
df_params_rest <- df_params %>% filter(program != "BEAST")

g <- ggplot() +
  my_theme +
  geom_histogram(data = df_params_beast, aes(x = value, y = after_stat(density), colour = label2), fill = "#999999", alpha = 0.5, bins = 50, position = "identity") +
  stat_density(data = df_params_rest, aes(x = value, colour = label2), geom = "line", position = "identity", trim = T) +
  scale_color_manual(values = c("#999999", "#009E73", "#0072B2", "#D55E00")) +
  facet_wrap(~param, scales = "free", labeller = human_readable_labeller) +
  labs(colour = NULL, fill = NULL, x = "Value", y = "Density")

reposition_legend(g, "center", panel = "panel-3-2")
```

\newpage

```{r}
df_params <- params %>%
  filter(model == "skygrid", engine != "torchtree") %>%
  mutate(label2 = factor(label2, levels = c("BEAST MCMC", "torchtree HMC", "torchtree MCMC", "torchtree ELBO", "torchtree KL(p||q)")))

df_params_beast <- df_params %>% filter(program == "BEAST")
df_params_rest <- df_params %>% filter(program != "BEAST")

# cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
cbbPalette <- c("#999999", "#009E73", "#E69F00", "#0072B2", "#D55E00")

g <- ggplot() +
  my_theme +
  geom_histogram(data = df_params_beast, aes(x = value, y = after_stat(density), colour = label2), fill = "#999999", alpha = 0.5, bins = 50, position = "identity") +
  # geom_density(data = df_params_rest, aes(x = value, colour = label2), alpha = 1, position = "identity") +
  stat_density(data = df_params_rest, aes(x = value, colour = label2), geom = "line", position = "identity", trim = T) +
  scale_color_manual(values = cbbPalette) +
  facet_wrap(~param, scales = "free", labeller = human_readable_labeller) +
  labs(colour = NULL, fill = NULL, x = "Value", y = "Density")

reposition_legend(g, "center", panel = "panel-3-2")
```

\newpage

## Densities with skygrid
```{r skygrid-probs}
df_probs <- probs %>%
  filter(model == "skygrid", engine != "torchtree")

ggplot(df_probs, aes(value, group = label, colour = label)) +
  my_theme +
  geom_density() +
  facet_wrap(~param, scales = "free", nrow = 2) +
  labs(colour = NULL, x = "Value", y = "Density")
```

\newpage

```{r}
kable(times)
```

\newpage

# Skyglide 

## Parameter densities with skyglide
```{r skyglide-params}
df_params <- params %>%
  filter(model == "skyglide", engine != "torchtree") %>%
  filter(label2 != "torchtree MCMC") %>%
  mutate(label2 = factor(label2, levels = c("BEAST MCMC", "torchtree HMC", "torchtree ELBO", "torchtree KL(p||q)")))

df_params_beast <- df_params %>% filter(program == "BEAST")
df_params_rest <- df_params %>% filter(program != "BEAST")

cbbPalette <- c("#999999", "#009E73", "#0072B2", "#D55E00")

g <- ggplot() +
  my_theme +
  geom_histogram(
    data = df_params_beast,
    aes(x = value, y = after_stat(density), colour = label2),
    fill = "#999999", alpha = 0.5, bins = 50, position = "identity"
  ) +
  # geom_density(data = df_params_rest, aes(x = value, colour = label), alpha = 1, position = "identity") +
  stat_density(data = df_params_rest, aes(x = value, colour = label2), geom = "line", position = "identity", trim = T) +
  scale_color_manual(values = cbbPalette) +
  facet_wrap(~param, scales = "free", labeller = human_readable_labeller) +
  labs(colour = NULL, fill = NULL, x = "Value", y = "Density")

reposition_legend(g, "center", panel = "panel-3-2")
```

\newpage

```{r}
df_params <- params %>%
  filter(model == "skyglide", engine != "torchtree") %>%
  mutate(label2 = factor(label2, levels = c("BEAST MCMC", "torchtree HMC", "torchtree MCMC", "torchtree ELBO", "torchtree KL(p||q)")))

df_params_beast <- df_params %>% filter(program == "BEAST")
df_params_rest <- df_params %>% filter(program != "BEAST")

cbbPalette <- c("#999999", "#009E73", "#E69F00", "#0072B2", "#D55E00")

g <- ggplot() +
  my_theme +
  geom_histogram(
    data = df_params_beast,
    aes(x = value, y = after_stat(density), colour = label2),
    fill = "#999999", alpha = 0.5, bins = 50, position = "identity"
  ) +
  # geom_density(data = df_params_rest, aes(x = value, colour = label), alpha = 1, position = "identity") +
  stat_density(data = df_params_rest, aes(x = value, colour = label2), geom = "line", position = "identity", trim = T) +
  scale_color_manual(values = cbbPalette) +
  facet_wrap(~param, scales = "free", labeller = human_readable_labeller) +
  labs(colour = NULL, fill = NULL, x = "Value", y = "Density")

reposition_legend(g, "center", panel = "panel-3-2")
```

\newpage

## Densities with skyglide
```{r skyglide-probs}
df_probs <- probs %>%
  filter(model == "skyglide", engine != "torchtree")

ggplot(df_probs, aes(value, group = label, colour = label)) +
  my_theme +
  geom_density() +
  facet_wrap(~param, scales = "free", nrow = 2) +
  labs(colour = NULL, x = "Value", y = "Density")
```

\newpage

# Skygrid and skyglide

## Just BEAST
```{r sky-beast}
df <- params %>%
  filter(program == "BEAST")

g <- ggplot(df, aes(value, colour = label)) +
  my_theme +
  geom_density() +
  facet_wrap(~param, scales = "free", labeller = human_readable_labeller) +
  labs(x = "Value", y = "Density", colour = NULL)

reposition_legend(g, "center", panel = "panel-3-2")
```
```

\newpage

## Parameter densities with skygrid and skyglide
```{r}
df <- params %>%
  filter(engine != "torchtree")

ggplot(df, aes(value, colour = label)) +
  my_theme +
  geom_density() +
  facet_wrap(~param, scales = "free") +
  labs(colour = NULL, x = "Value", y = "Density")
```

\newpage

## Densities with skygrid and skyglide
```{r}
df <- probs %>%
  filter(engine != "torchtree")

ggplot(df, aes(value, colour = label)) +
  my_theme +
  geom_density() +
  facet_wrap(~param, scales = "free") +
  labs(colour = NULL, x = "Value", y = "Density")
```

\newpage

## Skygrid and skyglide plots
```{r sky-plots}
cbbPalette <- c("#999999", "#009E73", "#0072B2", "#D55E00")
skys2 <- skys %>%
  filter(label2 != "torchtree MCMC", engine != "torchtree") %>%
  mutate(
    proghow = str_replace(label, "-skyg\\w+", ""),
    model = factor(model, levels = c("skygrid", "skyglide")),
    label2 = factor(label2, levels = c("BEAST MCMC", "torchtree HMC", "torchtree ELBO", "torchtree KL(p||q)"))
  )
skyplot(skys2, fill_by = "label2", facet = list(rows = "model", cols = "label2")) +
  theme(legend.position = "none", axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  scale_x_continuous(breaks = c(1700, 1800, 1900, 2000)) +
  scale_fill_manual(values = cbbPalette, name = "fill") +
  labs(x = "Year", y = "Effective population size")
```
```

\newpage

# Speed table
```{r}
infos <- NULL
for (coalescent in c("skyglide", "skygrid")) {
  path <- file.path("results", "HCV", "beast", coalescent)
  t <- parse_time(file.path(path, "beast.log"))
  info <- data.frame(
    dataset = "HCV", model = "GTR", engine = "BEAGLE", time = t,
    divergence = "", iters = 50000000, elbo = "",
    coalescent = coalescent, program = "BEAST"
  )
  infos <- rbind(infos, info)

  for (engine in c("physher", "torchtree")) {
    for (algorithm in c("hmc", "mcmc")) {
    path <- file.path("results", "HCV", "torchtree", "GTR", coalescent, algorithm, engine)
    if (file.exists(path)) {
      t <- parse_time(file.path(path, "torchtree.log"))
      info <- data.frame(
        dataset = "HCV", model = "GTR", engine = engine, time = t,
          divergence = toupper(algorithm), iters = 10000000, elbo = "",
          coalescent = coalescent, program = "torchtree"
      )
      infos <- rbind(infos, info)
    }
    }

    for (divergence in c("ELBO", "KLpq-10")) {
      path <- file.path("results", "HCV", "torchtree", "GTR", coalescent, divergence, engine)
      if (file.exists(path)) {
        info <- read.csv(file.path(path, "info.csv")) %>% mutate(program = "torchtree")
        infos <- rbind(infos, info)
      }
    }
  }
}
infos <- mutate(infos, "time (hours)" = time / 24, "hours/million" = time / iters * 1e6 / 24) %>%
  select(-dataset, -model) %>%
  mutate(engine = ifelse(engine == "torchtree", "AD", engine)) %>%
  select(program, engine, divergence, `time (hours)`, iters, `hours/million`, coalescent, elbo)
kable(infos)
```
