---
title: "SC2 Bayesian bridge"
author: "Mathieu Fourment"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output: 
  pdf_document:
    number_sections: true
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
library(ggplot2)
library(tidyr)
library(dplyr)
library(knitr)
library(data.table)
library(jsonlite)
library(stringr)
library(readr)
library(lemon)
knitr::opts_chunk$set(echo = TRUE, fig.path = "Figs/SC2/")
```

```{r include=FALSE}
source("R/utils.R")
```

The topology is fixed (iqtree + lsd2) and simple Bayesian bridge (no local scales) is used.

\newpage

```{r}
resultsdir <- file.path("results", "SC2")
freqs <- c(
  0.2988387179135696,
  0.18371883310279738,
  0.1958960436176954,
  0.32154640536593765
)

dlnorm2 <- function(x, mean, stdev) {
  mu <- log(mean / sqrt(1 + (stdev * stdev) / (mean * mean)))
  sigma <- sqrt(log(1 + (stdev * stdev) / (mean * mean)))
  dlnorm(x, mu, sigma)
}

# beast has some priors on fixed parameters that are not included in torchtree
extra_prior <- log(dexp(0, 0.3333333333333)) + # clock.scale.prior
  log(dlnorm(8.0E-4, log(8.0E-4), 1.17481)) + # clock.location.prior
  log(dlnorm2(1, 1, 1)) * (583 * 2 - 2) + # ratesPrior
  -6 * log(6) # subst rates sum to 6

cbbPalette <- c("#000000", "#0072B2", "#D55E00")
```


# GTR models
```{r}
df_beast_gtr <- fread(file.path(resultsdir, "beast", "GTR", "skyglide", "SC2_GTR.log"),
  sep = "\t", skip = "state"
)
df_beast_gtr <- df_beast_gtr[(nrow(df_beast_gtr) * 0.5):nrow(df_beast_gtr), ]

gtr_rates <- df_beast_gtr %>%
  select(starts_with("gtr.rates") | starts_with("frequencies")) %>%
  rename_with(~ str_replace(., "gtr.rates.rate", "")) %>%
  rename(
    freqA = "frequencies1", freqC = "frequencies2",
    freqG = "frequencies3", freqT = "frequencies4"
  ) %>%
  mutate(rateSum = select(., AC, AG, AT, CG, CT, GT) %>%
    rowSums()) %>%
  mutate_at(
    c("AC", "AG", "AT", "CG", "CT", "GT"),
    ~ . / rateSum
  ) %>%
  mutate(
    CA = AC * freqA, GA = AG * freqA, TA = AT * freqA,
    GC = CG * freqC, TC = CT * freqC, TG = GT * freqG
  ) %>%
  mutate(
    AC = AC * freqC, AG = AG * freqG, AT = AT * freqT,
    CG = CG * freqG, CT = CT * freqT, GT = GT * freqT
  ) %>%
  mutate(norm = ((AC + AG + AT) * freqs[1] + (CA + CG + CT) * freqs[2] +
    (GA + GC + GT) * freqs[3] + (TA + TC + TG) * freqs[4])) %>%
  mutate_at(
    c("AC", "AG", "AT", "CG", "CT", "GT", "CA", "GA", "TA", "GC", "TC", "TG"),
    ~ . / norm
  ) %>%
  select(!starts_with("freq"), -rateSum, -norm) %>%
  mutate(type = "BEAST", model = "GTR", label = "BEAST MCMC") %>%
  gather("param", "value", -type, -model, -label)


gtr_probs <- df_beast_gtr %>%
  select(
    any_of(c("likelihood", "skygrid", "prior"))
  ) %>%
  rename(
    coalescent = "skygrid"
  ) %>%
  mutate(type = "BEAST", model = "GTR", label = "BEAST MCMC") %>%
  gather("param", "value", -type, -model, -label)

gtr_other <- df_beast_gtr %>%
  select(
    all_of(c("alpha", "treeModel.rootHeight")),
    starts_with("gtr.rates")
  ) %>%
  rename_with(~ str_replace(., "gtr.rates.rate", "")) %>%
  mutate_at(c("AC", "AG", "AT", "CG", "CT", "GT"), ~ . / 6) %>%
  rename(
    rootHeight = "treeModel.rootHeight"
  ) %>%
  mutate(type = "BEAST", model = "GTR", label = "BEAST MCMC") %>%
  gather("param", "value", -type, -model, -label)

t_lookup_gtr <- c(
  AC = "substmodel.rates.0", AG = "substmodel.rates.1", AT = "substmodel.rates.2",
  CG = "substmodel.rates.3", CT = "substmodel.rates.4", GT = "substmodel.rates.5"
)

gtr_thetas <- df_beast_gtr %>%
  select(starts_with("skygrid.log"), skygrid.precision) %>%
  rename_with(~ str_replace(., "skygrid.", "")) %>%
  rename_with(~ str_replace(., "logPopSize", "log population size ")) %>%
  rename(all_of(c("GMRF precision" = "precision"))) %>%
  mutate(type = "BEAST", model = "GTR", label = "BEAST MCMC") %>%
  gather("param", "value", -type, -model, -label)

for (divergence in c("ELBO", "KLpq-10")) {
  d <- if (divergence == "KLpq-10") "KL(p||q)" else divergence
  df_t_gtr <- fread(
    file.path(resultsdir, "torchtree", "GTR", "skyglide", divergence, "torchtree", "samples.csv"),
    sep = "\t"
  )

  gtr_rates <- df_t_gtr %>%
    select(starts_with("substmodel.rates")) %>%
    rename(all_of(t_lookup_gtr)) %>%
    mutate(
      CA = AC * freqs[1], GA = AG * freqs[1], TA = AT * freqs[1],
      GC = CG * freqs[2], TC = CT * freqs[2], TG = GT * freqs[3]
    ) %>%
    mutate(
      AC = AC * freqs[2], AG = AG * freqs[3], AT = AT * freqs[4],
      CG = CG * freqs[3], CT = CT * freqs[4], GT = GT * freqs[4]
    ) %>%
    mutate(norm = ((AC + AG + AT) * freqs[1] + (CA + CG + CT) * freqs[2] +
      (GA + GC + GT) * freqs[3] + (TA + TC + TG) * freqs[4])) %>%
    mutate_at(
      c("AC", "AG", "AT", "CG", "CT", "GT", "CA", "GA", "TA", "GC", "TC", "TG"),
      ~ . / norm
    ) %>%
    select(-norm) %>%
    mutate(type = d, model = "GTR", label = paste("torchtree", d)) %>%
    gather("param", "value", -type, -model, -label) %>%
    bind_rows(gtr_rates)

  gtr_probs <- df_t_gtr %>%
    select(
      like, coalescent, variational, prior
    ) %>%
    rename(likelihood = like) %>%
    mutate(prior = prior + extra_prior) %>%
    mutate(type = d, model = "GTR", label = paste("torchtree", d)) %>%
    gather("param", "value", -type, -model, -label) %>%
    bind_rows(gtr_probs)

  gtr_other <- df_t_gtr %>%
    select(
      sitemodel.shape.0, tree.root_height.0,
      starts_with("substmodel.rates")
    ) %>%
    rename(all_of(t_lookup_gtr)) %>%
    rename(all_of(c(
      alpha = "sitemodel.shape.0",
      rootHeight = "tree.root_height.0"
    ))) %>%
    mutate(type = d, model = "GTR", label = paste("torchtree", d)) %>%
    gather("param", "value", -type, -model, -label) %>%
    bind_rows(gtr_other)

  gtr_thetas <- df_t_gtr %>%
    select(starts_with("coalescent.theta"), gmrf.precision.0) %>%
    mutate(across(starts_with("coalescent.theta"), log)) %>%
    rename(
      logPopSize1 = coalescent.theta.0, logPopSize2 = coalescent.theta.1,
      logPopSize3 = coalescent.theta.2, logPopSize4 = coalescent.theta.3,
      logPopSize5 = coalescent.theta.4
    ) %>%
    rename_with(~ str_replace(., "logPopSize", "log population size ")) %>%
    rename(all_of(c("GMRF precision" = "gmrf.precision.0"))) %>%
    mutate(type = d, model = "GTR", label = paste("torchtree", d)) %>%
    gather("param", "value", -type, -model, -label) %>%
    bind_rows(gtr_thetas)
}
```

\newpage

## Probs
```{r gtr-probs}
ggplot(gtr_probs, aes(value, colour = label)) +
  my_theme +
  geom_density() +
  facet_wrap(~param, scales = "free") +
  scale_color_manual(values = cbbPalette) +
  labs(colour = NULL, x = "Value", y = "Density")
```

\newpage

## Tree likelihood parameters
```{r gtr-params}
human_readable_labeller <- function(x) {
  human_readable_labels <- expression(
    AC = "A" %<->% "C relative rate",
    AG = "A" %<->% "G relative rate",
    AT = "A" %<->% "T relative rate",
    CG = "C" %<->% "G relative rate",
    CT = "C" %<->% "T relative rate",
    GT = "G" %<->% "T relative rate",
    rootHeight = "root height",
    alpha = "rate variation shape"
  )
  list(as.list(human_readable_labels)[x$param])
}

gtr_other$param <- factor(gtr_other$param,
  levels = c("AC", "AG", "AT", "CG", "CT", "GT", "alpha", "rootHeight")
)

g <- ggplot(gtr_other, aes(value, colour = label)) +
  my_theme +
  geom_density() +
  facet_wrap(~param, scales = "free", labeller = human_readable_labeller) +
  scale_color_manual(values = cbbPalette) +
  labs(colour = NULL, x = "Value", y = "Density")

reposition_legend(g, "center", panel = "panel-3-3")
```

\newpage

## Coalescent
```{r gtr-coalescent}
ggplot(gtr_thetas, aes(value, colour = label)) +
  my_theme +
  geom_density() +
  facet_wrap(~param, scales = "free") +
  scale_color_manual(values = cbbPalette) +
  labs(colour = NULL, x = "Value", y = "Density")
```

\newpage

```{r}
df <- df_beast_gtr %>%
  rename(
    rootHeight = "treeModel.rootHeight"
  ) %>%
  select(starts_with("skygrid.log"), rootHeight, state) %>%
  rename_with(~ str_replace(., "skygrid.", "")) %>%
  gather("param", "value", -state)

ggplot(df, aes(x = state, y = value)) +
  geom_line() +
  facet_wrap(~param, scales = "free")
```

```{r}
df <- df_beast_gtr %>%
  select(skygrid, likelihood, prior, state) %>%
  gather("param", "value", -state)

ggplot(df, aes(x = state, y = value)) +
  geom_line() +
  facet_wrap(~param, scales = "free", nrow = 2)
```

\newpage

# HKY-RE models
```{r}
df_beast <- fread(
  file.path(resultsdir, "beast", "HKY-RE", "skyglide", "SC2_MH.log"),
  sep = "\t",
  skip = "state"
)
df_beast <- df_beast[(nrow(df_beast) * 0.5):nrow(df_beast), ]
beast_coefs <- df_beast %>% select(starts_with("glmRandCoefficients"))

lookup <- sapply(1:12, function(x) {
  paste0("glmRandCoefficients", x)
})
names(lookup) <- c("AC", "AG", "AT", "CG", "CT", "GT", "CA", "GA", "TA", "GC", "TC", "TG")

coefs <- beast_coefs %>%
  rename(all_of(lookup)) %>%
  mutate(type = "BEAST", model = "HKY-RE", label = "BEAST MCMC") %>%
  gather("param", "value", -type, -model, -label)

hky_other <- df_beast %>%
  select(all_of(
    c(
      "randomEffects.globalScale",
      "alpha",
      "log.kappa",
      "treeModel.rootHeight",
      "skygrid.precision"
    )
  )) %>%
  mutate(kappa = exp(log.kappa)) %>%
  select(-log.kappa) %>%
  rename(all_of(
    c(
      "rate variation shape" = "alpha",
      "Bayesian bridge scale" = "randomEffects.globalScale",
      "root height" = "treeModel.rootHeight",
      "GMRF precision" = "skygrid.precision",
      "HKY kappa" = "kappa"
    )
  )) %>%
  mutate(type = "BEAST", model = "HKY-RE", label = "BEAST MCMC") %>%
  gather("param", "value", -type, -model, -label)

hky_probs <- df_beast %>%
  select(all_of(
    c(
      "likelihood",
      "skygrid"
    )
  )) %>%
  rename(all_of(
    c(coalescent = "skygrid")
  )) %>%
  mutate(type = "BEAST", model = "HKY-RE", label = "BEAST MCMC") %>%
  gather("param", "value", -type, -model, -label)

lookup <- setNames(
  c("log.kappa", sapply(1:12, function(x) {
    paste0("glmRandCoefficients", x)
  })),
  c("kappa", "AC", "AG", "AT", "CG", "CT", "GT", "CA", "GA", "TA", "GC", "TC", "TG")
)
hky_rates <- df_beast %>%
  select(starts_with("glmRandCoefficients") | log.kappa) %>%
  exp() %>%
  rename(all_of(lookup)) %>%
  # mutate(rateSum = select(., AC, AG, AT, CG, CT, GT, CA, GA, TA, GC, TC, TG) %>%
  #     rowSums()) %>%
  # mutate_at(
  #     c("AC", "AG", "AT", "CG", "CT", "GT", "CA", "GA", "TA", "GC", "TC", "TG"),
  #     ~ . / rateSum
  # ) %>%
  mutate(
    AC = AC * freqs[2], AG = AG * freqs[3] * kappa, AT = AT * freqs[4],
    CG = CG * freqs[3], CT = CT * freqs[4] * kappa, GT = GT * freqs[4]
  ) %>%
  mutate(
    CA = CA * freqs[1], GA = GA * freqs[1] * kappa, TA = TA * freqs[1],
    GC = GC * freqs[2], TC = TC * freqs[2] * kappa, TG = TG * freqs[3]
  ) %>%
  mutate(norm = ((AC + AG + AT) * freqs[1] + (CA + CG + CT) * freqs[2] +
    (GA + GC + GT) * freqs[3] + (TA + TC + TG) * freqs[4])) %>%
  mutate_at(
    c("AC", "AG", "AT", "CG", "CT", "GT", "CA", "GA", "TA", "GC", "TC", "TG"),
    ~ . / norm
  ) %>%
  select(-c(kappa, norm)) %>%
  mutate(type = "BEAST", model = "HKY-RE", label = "BEAST MCMC") %>%
  gather("param", "value", -type, -model, -label)

hky_thetas <- df_beast %>%
  select(starts_with("skygrid.log"), skygrid.precision) %>%
  rename_with(~ str_replace(., "skygrid.", "")) %>%
  mutate(type = "BEAST", model = "HKY-RE", label = "BEAST MCMC") %>%
  gather("param", "value", -type, -model, -label)

for (divergence in c("ELBO", "KLpq-10")) {
  d <- if (divergence == "KLpq-10") "KL(p||q)" else divergence
  df_t <- fread(file.path(resultsdir, "torchtree", "HKY-RE", "skyglide", divergence, "torchtree", "samples.csv"),
    sep = "\t"
  )
  t_coefs <- df_t %>% select(starts_with("substmodel.kappa.random"))

  lookup <- sapply(0:11, function(x) {
    paste0("substmodel.kappa.random.", x)
  })
  names(lookup) <- c("AC", "AG", "AT", "CG", "CT", "GT", "CA", "GA", "TA", "GC", "TC", "TG")

  coefs <- t_coefs %>%
    rename(all_of(lookup)) %>%
    mutate(type = d, model = "HKY-RE", label = paste("torchtree", d)) %>%
    gather("param", "value", -type, -model, -label) %>%
    bind_rows(coefs)

  hky_other <- df_t %>%
    select(
      sitemodel.shape.0,
      bayesian.bridge.scale.0,
      substmodel.kappa.unres.0,
      tree.root_height.0,
      gmrf.precision.0
    ) %>%
    mutate(kappa = exp(substmodel.kappa.unres.0)) %>%
    select(-substmodel.kappa.unres.0) %>%
    rename(all_of(
      c(
        "Bayesian bridge scale" = "bayesian.bridge.scale.0",
        "rate variation shape" = "sitemodel.shape.0",
        "HKY kappa" = "kappa",
        "root height" = "tree.root_height.0",
        "GMRF precision" = "gmrf.precision.0"
      )
    )) %>%
    mutate(type = d, model = "HKY-RE", label = paste("torchtree", d)) %>%
    gather("param", "value", -type, -model, -label) %>%
    bind_rows(hky_other)

  hky_probs <- df_t %>%
    select(
      like,
      coalescent
    ) %>%
    rename(all_of(
      c(
        likelihood = "like"
      )
    )) %>%
    mutate(type = d, model = "HKY-RE", label = paste("torchtree", d)) %>%
    gather("param", "value", -type, -model, -label) %>%
    bind_rows(hky_probs)

  lookup <- sapply(0:11, function(x) {
    paste0("substmodel.kappa.", x)
  })
  names(lookup) <- c("AC", "AG", "AT", "CG", "CT", "GT", "CA", "GA", "TA", "GC", "TC", "TG")

  hky_rates <- df_t %>%
    select(matches("substmodel.kappa.\\d+")) %>%
    rename(all_of(lookup)) %>%
    # mutate(rateSum = select(., AC, AG, AT, CG, CT, GT, CA, GA, TA, GC, TC, TG) %>%
    #     rowSums()) %>%
    # mutate_at(
    #     c("AC", "AG", "AT", "CG", "CT", "GT", "CA", "GA", "TA", "GC", "TC", "TG"),
    #     ~ . / rateSum
    # ) %>%
    mutate(
      AC = AC * freqs[2], AG = AG * freqs[3], AT = AT * freqs[4],
      CG = CG * freqs[3], CT = CT * freqs[4], GT = GT * freqs[4]
    ) %>%
    mutate(
      CA = CA * freqs[1], GA = GA * freqs[1], TA = TA * freqs[1],
      GC = GC * freqs[2], TC = TC * freqs[2], TG = TG * freqs[3]
    ) %>%
    mutate(norm = ((AC + AG + AT) * freqs[1] + (CA + CG + CT) * freqs[2] +
      (GA + GC + GT) * freqs[3] + (TA + TC + TG) * freqs[4])) %>%
    mutate_at(
      c("AC", "AG", "AT", "CG", "CT", "GT", "CA", "GA", "TA", "GC", "TC", "TG"),
      ~ . / norm
    ) %>%
    select(-norm) %>%
    mutate(type = d, model = "HKY-RE", label = paste("torchtree", d)) %>%
    gather("param", "value", -type, -model, -label) %>%
    bind_rows(hky_rates)

  hky_thetas <- df_t %>%
    select(starts_with("coalescent.theta"), gmrf.precision.0) %>%
    mutate(across(starts_with("coalescent.theta"), log)) %>%
    rename(
      logPopSize1 = coalescent.theta.0, logPopSize2 = coalescent.theta.1,
      logPopSize3 = coalescent.theta.2, logPopSize4 = coalescent.theta.3,
      logPopSize5 = coalescent.theta.4
    ) %>%
    rename_with(~ str_replace(., "logPopSize", "log population size ")) %>%
    rename(all_of(c("GMRF precision" = "gmrf.precision.0"))) %>%
    mutate(type = d, model = "HKY-RE", label = paste("torchtree", d)) %>%
    gather("param", "value", -type, -model, -label) %>%
    bind_rows(gtr_thetas)
}
```

\newpage

## Random effects ($\epsilon$)
```{r hkyre-coefs}
human_readable_labeller <- function(x) {
  human_readable_labels <- expression(
    AC = "A" %->% "C", CA = "C" %->% "A",
    AG = "A" %->% "G", GA = "G" %->% "A",
    AT = "A" %->% "T", TA = "T" %->% "A",
    CG = "C" %->% "G", GC = "G" %->% "C",
    CT = "C" %->% "T", TC = "T" %->% "C",
    GT = "G" %->% "T", TG = "T" %->% "G"
  )
  list(as.list(human_readable_labels)[x$param])
}

ggplot(coefs, aes(value, colour = label)) +
  my_theme +
  geom_density(bw = 0.2) +
  geom_vline(xintercept = 0) +
  facet_wrap(~param, scales = "free", labeller = human_readable_labeller) +
  scale_color_manual(values = cbbPalette) +
  labs(colour = NULL, x = "Value", y = "Density")
```

\newpage

## Probs
```{r hkre-probs}
ggplot(hky_probs, aes(value, colour = label)) +
  my_theme +
  geom_density() +
  facet_wrap(~param, scales = "free") +
  scale_color_manual(values = cbbPalette) +
  labs(colour = NULL, x = "Value", y = "Density")
```

\newpage

## Coalescent
```{r hkyre-coalescent}
ggplot(hky_thetas, aes(value, colour = label)) +
  my_theme +
  geom_density() +
  facet_wrap(~param, scales = "free") +
  scale_color_manual(values = cbbPalette) +
  labs(colour = NULL, x = "Value", y = "Density")
```

\newpage
## Tree likelihood parameters
```{r hkyre-params}
human_readable_labeller <- function(x) {
  human_readable_labels <- x$param
  human_readable_labels[which(x$param == "HKY kappa")] <- expression(paste("HKY ", kappa))
  list(as.list(human_readable_labels))
}

g <- ggplot(hky_other, aes(value, colour = label)) +
  my_theme +
  geom_density() +
  facet_wrap(~param, scales = "free", labeller = human_readable_labeller) +
  scale_color_manual(values = cbbPalette) +
  labs(colour = NULL, x = "Value", y = "Density")

reposition_legend(g, "center", panel = "panel-3-2")
```


\newpage

# Normalized rates

```{r}
hky_rates$param <- factor(hky_rates$param,
  levels = c("AC", "CA", "AG", "GA", "AT", "TA", "CG", "GC", "CT", "TC", "GT", "TG")
)
df <- gtr_rates %>%
  bind_rows(hky_rates) %>%
  mutate(what = paste(model, type, sep = "-"))
ggplot(df, aes(value, param, color = what)) +
  geom_boxplot(outlier.shape = NA) +
  coord_flip() +
  scale_x_continuous(trans = "log2", breaks = c(0.01, 0.05, 0.2, 0.5, 2.0)) +
  xlab("Normalized rate") +
  ylab("Substitution type")
```

\newpage

```{r nice-plot}
temp <- data.frame(
  param = c("AC", "CA", "AG", "GA", "AT", "TA", "CG", "GC", "CT", "TC", "GT", "TG"),
  index = c(1, 1, 2, 2, 3, 3, 4, 4, 5, 5, 6, 6)
)
df <- gtr_rates %>%
  bind_rows(hky_rates) %>%
  filter(type != "KL(p||q)") %>%
  mutate(what = paste(model, type, sep = "-"))
df <- df %>% left_join(temp, by = join_by(param))

df[c("i", "j")] <- str_split_fixed(df$param, "", 2)
# generate warning in pdf and replace arrow with ..
df$formatted <- paste(df$i, "\u2192", df$j)

# does not work
# df$formatted <- sapply(1:nrow(df), function(ii) {
#   bquote(.(df$i[ii]) %->% .(df$j[ii]))
# })

ggplot(df, aes(value, param, color = what, fill = what)) +
  theme_bw() +
  theme(
    legend.position = "top",
    legend.title = element_blank(),
    strip.background = element_blank(),
    strip.text.x = element_blank()
  ) +
  geom_boxplot(outlier.shape = NA, alpha = 0.5) +
  coord_flip() +
  scale_x_continuous(
    trans = "log2",
    breaks = c(0.01, 0.02, 0.05, 0.1, 0.2, 0.5, 1.0, 2.0)
  ) +
  xlab("Normalized rate") +
  ylab("Substitution type") +
  # scale_colour_manual(values = c("#E69F00", "#009E73"), name = "") +
  # scale_fill_manual(values = c("#E69F00", "#009E73"), name = "") +
  facet_grid(. ~ index, drop = TRUE, scale = "free")
```

\newpage

# Bayes factors

Let $\phi = \tau^{-\alpha}$ and $\phi \sim Gamma(1, rate=0.5)$.
Using the change of variable formula the pdf of $\tau$ is

$$p(\tau) = p_\phi(\phi) |d\phi/d\tau| = p_\phi(\tau^{-\alpha}) |-\alpha \tau^{-\alpha-1}|$$

The prior density at $\Delta_{ij} = \epsilon_{ji} - \epsilon_{ij}=0$ given $\tau$ is

$$p(\Delta_{ij}=0|\tau) = \frac{\alpha}{2^{1+1/\alpha} \tau \Gamma(1/\alpha)}$$

Integrating $\tau$ we get

$$p(\Delta_{ij}=0) = \int_0^\infty p(\Delta_{ij}=0|\tau) p(\tau) d\tau.$$

The Bayes Factor in favor of $\Delta_{ij} = 0$ (Model 0, against Model 1 where $\Delta_{ij}$ is a free parameter) is the ratio of the posterior density to the prior density at $\Delta_{ij} = 0$,

$$BF_{01} = \frac{p(\Delta_{ij}=0 | y)}{p(\Delta_{ij}=0)} \approx \frac{q(\Delta_{ij}=0)}{p(\Delta_{ij}=0)},$$

where $q$ is a normal distribution with mean $\mu_{ji}-\mu_{ij}$ and variance $\sigma_{ij}^2 + \sigma_{ji}^2$.
$\mu_{ij}$ and $\sigma_{ij}$ are the mean and standard deviation of the normal variational approximation of $\epsilon_{ij}$.

```{r}
p <- file.path(resultsdir, "torchtree", "HKY-RE", "skyglide", "ELBO", "torchtree")
if (!file.exists(file.path(p, "checkpoints"))) {
  untar(file.path(p, "checkpoints.tar.gz"), exdir = file.path(p, "checkpoints"))
}
j <- fromJSON(file.path(p, "checkpoints", "checkpoint-1000000.json"))
loc <- j$tensor[[match("var.Normal.substmodel.kappa.random.loc", j$id)]]
log_scale <- j$tensor[[match("var.Normal.substmodel.kappa.random.scale.unres", j$id)]]
scale <- exp(log_scale)

post0 <- dnorm(0., loc[7:12] - loc[1:6], sd = sqrt(scale[1:6]**2 + scale[7:12]**2))

alpha <- 0.25
prior_tau_0 <- function(tau) {
  alpha / (2.**(1. + 1. / alpha) * tau * gamma(1. / alpha)) *
    dgamma(tau**(-alpha), 1.0, rate = 0.5) * alpha * tau**(-alpha - 1)
}
prior0 <- integrate(prior_tau_0, lower = 0, upper = Inf)
df <- data.frame(
  Substitution = c("A<->C", "A<->G", "A<->T", "C<->G", "C<->T", "G<->T"),
  BF01 = post0 / prior0$value,
  BF10 = prior0$value / post0,
  logBF10 = log10(prior0$value / post0),
  posterior = post0,
  prior = prior0$value
)
kable(df)
```

```{r}
coeffs <- df_beast %>% select(starts_with("glmRandCoefficients"))
locs <- coeffs[, glmRandCoefficients1:glmRandCoefficients6] - coeffs[, glmRandCoefficients7:glmRandCoefficients12]
beast_post0 <- numeric(ncol(locs))
for (i in 1:ncol(locs)) {
  values <- locs[[paste0("glmRandCoefficients", i)]]
  # d = density(values)
  # beast_post0[i] = approx(d$x,d$y,xout=0.0)$y
  beast_post0[i] <- dnorm(0.0, mean(-values), sd(-values))
}
```

```{r}
temp <- locs %>% gather("param", "value")
ggplot(temp, aes(value, group = param)) +
  geom_density() +
  facet_wrap(~param, scales = "free")
```

```{r}
df <- data.frame(
  Substitution = c("A<->C", "A<->G", "A<->T", "C<->G", "C<->T", "G<->T"),
  BF01 = beast_post0 / prior0$value,
  BF10 = prior0$value / beast_post0,
  logBF10 = log10(prior0$value / beast_post0),
  posterior = beast_post0,
  prior = prior0$value
)
kable(df)
```

```{r}
df <- data.frame(
  Substitution = c("A<->C", "A<->G", "A<->T", "C<->G", "C<->T", "G<->T"),
  BEAST = log10(prior0$value / beast_post0),
  torchtree = log10(prior0$value / post0)
)
kable(df)
```

```{r, eval=FALSE,include=FALSE}
plot(seq(0, 1000000, 1), prior.tau.0(seq(0, 1000000, 1)))
```

\newpage

# Convergence
```{r}
use_budget <- T

p <- file.path("results", "SC2", "beast", "HKY-RE", "skyglide")
beast_time <- parse_time(file.path(p, "beast.log"))

df <- fread(
  file.path(p, "SC2_MH.log"),
  sep = "\t",
  skip = "#",
  data.table = FALSE
)

lookup <- c(
  sapply(1:5, function(x) {
    paste0("skygrid.logPopSize", x)
  }),
  sapply(1:12, function(x) {
    paste0("glmRandCoefficients", x)
  }),
  "treeModel.rootHeight",
  "skygrid.precision",
  "alpha",
  "randomEffects.globalScale"
)
names(lookup) <- c(
  sapply(1:5, function(x) {
    paste0("population size ", x)
  }),
  c("reAC", "reAG", "reAT", "reCG", "reCT", "reGT", "reCA", "reGA", "reTA", "reGC", "reTC", "reTG"),
  "root height", "GMRF precision", "rate variation shape", "Bayesian bridge scale"
)

params_id <- c(
  "root height",
  "kappa",
  "rate variation shape",
  "GMRF precision",
  "Bayesian bridge scale"
)
df <- df %>%
  mutate(kappa = exp(log.kappa)) %>%
  rename(all_of(lookup)) %>%
  select(state, all_of(params_id), contains("population"), starts_with("re"))

vars <- cvs <- means <- means2 <- df[FALSE, ]
var0 <- summarize_all(df, var)
sd0 <- sqrt(var0)
mean0 <- summarize_all(df, mean)
cv0 <- sd0 / mean0

every <- 500
replicates <- nrow(df) / every
state_count <- df$state[length(df$state)]

for (i in 1:replicates) {
  df_i <- df[1:(i * every), ]
  df2_var <- summarize_all(df_i, var)
  df2_mean <- summarize_all(df_i, mean)
  df2_sd <- sqrt(df2_var)
  df2_cv <- df2_sd / df2_mean
  vars <- bind_rows(vars, (df2_var - var0)**2)
  cvs <- bind_rows(cvs, (df2_cv - cv0)**2)
  means <- bind_rows(means, (df2_mean - mean0)**2)
  means2 <- bind_rows(means2, df2_mean)
}

vars$state <- seq(1, (replicates - 1) * every, every) * 500
cvs$state <- vars$state
means$state <- vars$state
means2$state <- vars$state

vars <- vars %>%
  mutate(
    time = beast_time / state_count * state, program = "BEAST",
    label = "BEAST MCMC", divergence = NA
  )
cvs <- cvs %>%
  mutate(
    time = beast_time / state_count * state, program = "BEAST",
    label = "BEAST MCMC", divergence = NA
  )
means <- means %>%
  mutate(
    time = beast_time / state_count * state, program = "BEAST",
    label = "BEAST MCMC", divergence = NA
  )
means2 <- means2 %>%
  mutate(
    time = beast_time / state_count * state, program = "BEAST",
    label = "BEAST MCMC", divergence = NA
  )
```

```{r}
lookup <- c(sapply(0:4, function(x) {
  paste0("coalescent.theta.", x)
}), sapply(0:11, function(x) {
  paste0("substmodel.kappa.random.", x)
}))
names(lookup) <- c(
  sapply(1:5, function(x) {
    paste0("population size ", x)
  }),
  c(
    "reAC", "reAG", "reAT", "reCG", "reCT", "reGT",
    "reCA", "reGA", "reTA", "reGC", "reTC", "reTG"
  )
)

lookup <- c(
  lookup,
  "root height" = "tree.root_height.0",
  "GMRF precision" = "gmrf.precision.0",
  log.kappa = "substmodel.kappa.unres.0",
  "rate variation shape" = "sitemodel.shape.0",
  likelihood = "like",
  "Bayesian bridge scale" = "bayesian.bridge.scale.0"
)

for (divergence in c("ELBO", "KLpq-10")) {
  p <- file.path(
    "results",
    "SC2",
    "torchtree",
    "HKY-RE",
    "skyglide",
    divergence,
    "torchtree"
  )
  v_log <- read_lines(file.path(p, "torchtree.log"))
  time <- as.numeric(v_log[length(v_log) - 2] %>% str_split_i("\t", 2) %>% str_split_i("m", 1))

  df_m <- fread(file.path(p, "means.csv"),
    sep = ",",
    data.table = FALSE
  )
  df_v <- fread(file.path(p, "variances.csv"),
    sep = ",",
    data.table = FALSE
  )

  params_id <- c(
    "root height",
    "kappa",
    "rate variation shape",
    "GMRF precision",
    "Bayesian bridge scale"
  )

  df_means2 <- df_m %>%
    rename(state = sample) %>%
    rename(any_of(lookup)) %>%
    mutate(kappa = exp(log.kappa)) %>%
    select(state, all_of(params_id), starts_with("population"), starts_with("re"))

  df_m <- df_m %>%
    rename(state = sample) %>%
    rename(any_of(lookup)) %>%
    mutate("kappa" = exp(log.kappa)) %>%
    select(state, all_of(params_id), starts_with("population"), starts_with("re")) %>%
    mutate(across(c(all_of(params_id), starts_with("population"), starts_with("re")), cummean))
  # mutate(across(everything(), cummean))

  df_v <- df_v %>%
    rename(state = sample) %>%
    rename(any_of(lookup)) %>%
    mutate(kappa = exp(log.kappa)) %>%
    select(state, all_of(params_id), starts_with("population"), starts_with("re")) %>%
    mutate(across(c(all_of(params_id), starts_with("population"), starts_with("re")), cummean))
  # mutate(across(everything(), cummean))

  d <- if (divergence == "KLpq-10") {
    "KL(p||q)"
  } else {
    divergence
  }
  iter <- as.numeric(df_m$state)

  # beast and torchtree would have run for the same amount of time (i.e. budget)
  if (use_budget) {
    total_iter <- iter[length(iter)]
    index <- which.min(abs(time / total_iter * df_v$state - beast_time))
    df_m <- df_m[1:index, ]
    df_v <- df_v[1:index, ]
    iter <- as.numeric(df_m$state)
    time <- time / total_iter * iter[length(iter)]
    df_means2 <- df_means2[1:index, ]
  }

  means2 <- df_means2 %>%
    mutate(
      program = "torchtree",
      divergence = d,
      label = paste("torchtree", d),
      time = time / iter[length(iter)] * state
    ) %>%
    bind_rows(means2)

  df_m2 <- sweep(df_m, 2, as.numeric(df_m[nrow(df_m), ]), "-")**2
  df_m2 <- df_m2 %>% mutate(
    state = iter,
    program = "torchtree",
    divergence = d,
    label = paste("torchtree", d),
    time = time / iter[length(iter)] * state
  )
  df_v2 <- sweep(df_v, 2, as.numeric(df_v[nrow(df_v), ]), "-")**2
  df_v2 <- df_v2 %>% mutate(
    state = iter,
    program = "torchtree",
    divergence = d,
    label = paste("torchtree", d),
    time = time / iter[length(iter)] * state
  )

  df_cv2 <- sweep(sqrt(df_v) / df_m, 2, as.numeric(sqrt(df_v[nrow(df_v), ]) /
    df_m[nrow(df_m), ]), "-")**2
  df_cv2 <- df_cv2 %>% mutate(
    state = iter,
    program = "torchtree",
    divergence = d,
    label = paste("torchtree", d),
    time = time / iter[length(iter)] * state
  )

  means <- rbind(means, df_m2)
  vars <- rbind(vars, df_v2)
  cvs <- rbind(cvs, df_cv2)
}
```

\newpage

```{r hkyre-means-params}
df2 <- means2 %>%
  # filter(time < 2000) %>%
  select(all_of(c(params_id, "state", "program", "time", "divergence", "label"))) %>%
  gather("param", "mean", -state, -time, -program, -divergence, -label)

human_readable_labeller <- function(x) {
  human_readable_labels <- x$param
  human_readable_labels[which(x$param == "kappa")] <- expression(paste("HKY ", kappa))
  list(as.list(human_readable_labels))
}

g <- ggplot(df2, aes(x = time, y = mean, color = label)) +
  my_theme +
  geom_line() +
  scale_color_manual(values = cbbPalette) +
  facet_wrap(~param, scales = "free", labeller = human_readable_labeller)

reposition_legend(g, "center", panel = "panel-3-2")
```

\newpage

```{r hkyre-cv-params}
df2 <- cvs %>%
  # filter(time < 2000) %>%
  select(all_of(c(params_id, "state", "program", "time", "divergence", "label"))) %>%
  gather("param", "CV", -state, -time, -program, -divergence, -label)

human_readable_labeller <- function(x) {
  human_readable_labels <- x$param
  human_readable_labels[which(x$param == "kappa")] <- expression(paste("HKY ", kappa))
  list(as.list(human_readable_labels))
}

g <- ggplot(df2, aes(x = time, y = CV, color = label)) +
  my_theme +
  geom_line() +
  # scale_y_continuous(trans = "log10") +
  facet_wrap(~param, scales = "free", labeller = human_readable_labeller) +
  scale_color_manual(values = cbbPalette) +
  labs(x = "Time (minutes)", y = bquote((CV[i] - CV[M])^2), color = NULL)
# labs(x = "Time (minutes)", y = bquote((sum(CV[j], j==1,i)/i - bar(CV)[M])^2), color = NULL)

reposition_legend(g, "center", panel = "panel-3-2")
```

\newpage

```{r hkyre-cv-coalescent}
df2 <- cvs %>%
  # filter(time < 2000) %>%
  select(contains("pop"), state, time, program, divergence, label) %>%
  gather("param", "CV", -state, -time, -program, -divergence, -label)

g <- ggplot(df2, aes(x = time, y = CV, color = label)) +
  my_theme +
  geom_line() +
  facet_wrap(~param, scales = "free") +
  scale_color_manual(values = cbbPalette) +
  labs(x = "Time (minutes)", y = bquote((CV[i] - CV[M])^2), color = NULL)

reposition_legend(g, "center", panel = "panel-3-2")
```

\newpage

```{r hkyre-cv-coefs}
df2 <- cvs %>%
  # filter(time < 2000) %>%
  select(starts_with("re"), all_of(c("state", "program", "time", "divergence", "label"))) %>%
  gather("param", "CV", -state, -time, -program, -divergence, -label) %>%
  filter(!is.nan(CV)) %>%
  mutate(param = factor(param, levels = c(
    "reAC", "reAG", "reAT", "reCG", "reCT", "reGT",
    "reCA", "reGA", "reTA", "reGC", "reTC", "reTG"
  )))

human_readable_labeller <- function(x) {
  human_readable_labels <- expression(
    reAC = "A" %->% "C", reCA = "C" %->% "A",
    reAG = "A" %->% "G", reGA = "G" %->% "A",
    reAT = "A" %->% "T", reTA = "T" %->% "A",
    reCG = "C" %->% "G", reGC = "G" %->% "C",
    reCT = "C" %->% "T", reTC = "T" %->% "C",
    reGT = "G" %->% "T", reTG = "T" %->% "G"
  )
  list(as.list(human_readable_labels)[x$param])
}

ggplot(df2, aes(x = time, y = CV, color = label)) +
  my_theme +
  geom_line() +
  scale_y_continuous(trans = "log10") +
  facet_wrap(~param, scales = "free", labeller = human_readable_labeller) +
  scale_color_manual(values = cbbPalette) +
  labs(x = "Time (minutes)", y = bquote((CV[i] - CV[M])^2), color = NULL)
```

\newpage

# VB traces
```{r}
df_elbos <- NULL
for (model in c("GTR", "HKY-RE")) {
  df <- fread(
    file.path(resultsdir, "torchtree", model, "skyglide", "ELBO", "torchtree", "elbo.csv"),
    sep = ","
  )
  df <- df %>%
    mutate(model = model) %>%
    rename(ELBO = "elbos", iteration = iters)
  df_elbos <- rbind(df_elbos, df)
}
```

```{r}
ggplot(df_elbos, aes(x = iteration, y = ELBO, color = model)) +
  geom_line()
```

```{r}
ggplot(df_elbos %>% filter(iteration > 100000), aes(x = iteration, y = ELBO, color = model)) +
  geom_line()
```

\newpage

# Speed table
```{r}
infos <- NULL
for (model in c("HKY-RE", "GTR")) {
  path <- file.path("results", "SC2", "beast", model, "skyglide")
  t <- parse_time(file.path(path, "beast.log"))
  info <- data.frame(
    dataset = "SC2", model = model, engine = "BEAST",
    time = t, divergence = "", iters = 50000000, elbo = "",
    coalescent = "skyglide"
  )
  infos <- rbind(infos, info)

  for (divergence in c("ELBO", "KLpq-10")) {
    for (engine in c("physher", "torchtree")) {
      path <- file.path("results", "SC2", "torchtree", model, "skyglide", divergence, engine)
      if (file.exists(path)) {
        info <- read.csv(file.path(path, "info.csv"))
        infos <- rbind(infos, info)
      }
    }
  }
}
infos <- mutate(infos, "time (hours)" = time / 24, "hours/million" = time / iters * 1e6 / 24) %>%
  select(-dataset, -coalescent) %>%
  mutate(program = ifelse(engine == "BEAST", "BEAST", "torchtree")) %>%
  mutate(engine = ifelse(engine == "BEAST", "MCMC", engine)) %>%
  mutate(engine = ifelse(engine == "torchtree", "AD", engine)) %>%
  select(program, model, engine, divergence, `time (hours)`, iters, `hours/million`, elbo)
kable(infos)
```
