---
title: "SC2 Bayesian bridge"
author: "Mathieu Fourment"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output: 
  pdf_document:
    number_sections: true
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
library(ggplot2)
library(tidyr)
library(dplyr)
library(knitr)
library(data.table)
library(jsonlite)
library(stringr)
knitr::opts_chunk$set(echo = TRUE, fig.path = "Figs/")
```


The topology is fixed (iqtree + lsd2) and simple Bayesian bridge (no local scales) is used.

\newpage

```{r}
resultsdir = file.path("results", "SC2")
freqs = c(0.2988387179135696,
          0.18371883310279738,
          0.1958960436176954,
          0.32154640536593765)
```


# GTR models
```{r}
df.beast.gtr = fread(file.path(resultsdir, "beast", "GTR", "skyglide", "SC2_GTR.log"),
    sep = "\t", skip = "state"
)
df.beast.gtr = df.beast.gtr[(nrow(df.beast.gtr) * 0.5):nrow(df.beast.gtr), ]

df.beast.gtr.rates = df.beast.gtr %>%
  select(starts_with("gtr.rates") | starts_with("frequencies")) %>%
  rename_with(~ str_replace(., "gtr.rates.rate", "")) %>%
  rename(
    freqA = "frequencies1", freqC = "frequencies2",
    freqG = "frequencies3", freqT = "frequencies4"
  ) %>%
  mutate(rateSum = select(., AC, AG, AT, CG, CT, GT) %>%
    rowSums()) %>%
  mutate_at(
    c("AC", "AG", "AT", "CG", "CT", "GT"),
    ~ . / rateSum
  ) %>%
  mutate(
    CA = AC * freqA, GA = AG * freqA, TA = AT * freqA,
    GC = CG * freqC, TC = CT * freqC, TG = GT * freqG
  ) %>%
  mutate(
    AC = AC * freqC, AG = AG * freqG, AT = AT * freqT,
    CG = CG * freqG, CT = CT * freqT, GT = GT * freqT
  ) %>%
  mutate(norm = ((AC + AG + AT) * freqs[1] + (CA + CG + CT) * freqs[2] +
    (GA + GC + GT) * freqs[3] + (TA + TC + TG) * freqs[4])) %>%
  mutate_at(
    c("AC", "AG", "AT", "CG", "CT", "GT", "CA", "GA", "TA", "GC", "TC", "TG"),
    ~ . / norm
  ) %>%
  select(!starts_with("freq"), -rateSum, -norm) %>%
  mutate(type = "GTR-beast") %>%
  gather("param", "value", -type)

df.t.gtr = fread(
  #file.path(resultsdir, "torchtree", "GTR", "skyglide", "ELBO", "samples.csv"),
  file.path(resultsdir, "torchtree", "GTR2", "ELBO", "samples.csv"),
  sep = "\t")

t.lookup.gtr = c(
  AC = "substmodel.rates.0", AG = "substmodel.rates.1", AT = "substmodel.rates.2",
  CG = "substmodel.rates.3", CT = "substmodel.rates.4", GT = "substmodel.rates.5"
)

df.t.gtr.rates = df.t.gtr %>%
  select(starts_with("substmodel.rates")) %>%
  rename(all_of(t.lookup.gtr)) %>%
  mutate(
    CA = AC * freqs[1], GA = AG * freqs[1], TA = AT * freqs[1],
    GC = CG * freqs[2], TC = CT * freqs[2], TG = GT * freqs[3]
  ) %>%
  mutate(
    AC = AC * freqs[2], AG = AG * freqs[3], AT = AT * freqs[4],
    CG = CG * freqs[3], CT = CT * freqs[4], GT = GT * freqs[4]
  ) %>%
  mutate(norm = ((AC + AG + AT) * freqs[1] + (CA + CG + CT) * freqs[2] +
    (GA + GC + GT) * freqs[3] + (TA + TC + TG) * freqs[4])) %>%
  mutate_at(
    c("AC", "AG", "AT", "CG", "CT", "GT", "CA", "GA", "TA", "GC", "TC", "TG"),
    ~ . / norm
  ) %>%
  select(-norm) %>%
    mutate(type = "GTR") %>%
    gather("param", "value", -type)

beast.gtr.other = df.beast.gtr %>%
  select(
    all_of(c("likelihood", "alpha", "treeModel.rootHeight")),
    starts_with("gtr.rates")
  ) %>%
  rename_with(~ str_replace(., "gtr.rates.rate", "")) %>%
  mutate_at(c("AC", "AG", "AT", "CG", "CT", "GT"), ~ . / 6) %>%
  rename(
    rootHeight = "treeModel.rootHeight"
  ) %>%
  mutate(type = "beast") %>%
  gather("param", "value", -type)

t.gtr.other = df.t.gtr %>%
  select(
    like, sitemodel.shape.0, tree.root_height.0,
    starts_with("substmodel.rates")
    ) %>%
  rename(all_of(t.lookup.gtr)) %>%
  rename(all_of(c(
    likelihood = "like", alpha = "sitemodel.shape.0",
    rootHeight = "tree.root_height.0"
  ))) %>%
  mutate(type = "torchtree") %>%
  gather("param", "value", -type)

gtr.other = rbind(beast.gtr.other, t.gtr.other)
```

```{r}
ggplot(gtr.other, aes(value, group = type, colour = type)) +
  geom_density() +
  facet_wrap( ~ param, scales = "free")
```

\newpage

# HKY-RE models
```{r}
df.beast = fread(
  file.path(resultsdir, "beast", "HKY-RE", "skyglide", "SC2_MH.log"),
  sep = "\t",
  skip = "state"
)
df.beast = df.beast[(nrow(df.beast) * 0.5):nrow(df.beast),]
beast.coefs = df.beast %>% select(starts_with("glmRandCoefficients"))

lookup = sapply(1:12, function(x) {
  paste0("glmRandCoefficients", x)
})
names(lookup) = c("AC", "AG", "AT", "CG", "CT", "GT", "CA", "GA", "TA", "GC", "TC", "TG")

beast.coefs = beast.coefs %>%
  rename(all_of(lookup)) %>%
  mutate(type = "BEAST") %>%
  gather("param", "value",-type)

df.t = fread(file.path(resultsdir, "torchtree", "HKY-RE", "skyglide", "ELBO", "samples.csv"),
             sep = "\t")
t.coefs = df.t %>% select(starts_with("substmodel.kappa.random"))

lookup = sapply(0:11, function(x) {
  paste0("substmodel.kappa.random.", x)
})
names(lookup) = c("AC", "AG", "AT", "CG", "CT", "GT", "CA", "GA", "TA", "GC", "TC", "TG")

t.coefs = t.coefs %>%
  rename(all_of(lookup)) %>%
  mutate(type = "torchtree") %>%
  gather("param", "value",-type)

coefs = rbind(t.coefs, beast.coefs)
```

\newpage

## Random effects ($\epsilon$)
```{r}
ggplot(coefs, aes(value, group = type, colour = type)) +
  geom_density() +
  geom_vline(xintercept = 0) +
  facet_wrap( ~ param, scales = "free")
```


\newpage
## Other parameters

```{r}
beast.other = df.beast %>%
  select(all_of(
    c(
      "likelihood",
      "randomEffects.globalScale",
      "alpha",
      "log.kappa",
      "treeModel.rootHeight"
    )
  )) %>%
  rename(all_of(
    c(globalScale = "randomEffects.globalScale", rootHeight = "treeModel.rootHeight")
  )) %>%
  mutate(type = "beast") %>%
  gather("param", "value",-type)

t.bb.other = df.t %>%
  select(
    like,
    sitemodel.shape.0,
    bayesian.bridge.scale.0,
    substmodel.kappa.unres.0,
    tree.root_height.0
  ) %>%
  rename(all_of(
    c(
      likelihood = "like",
      globalScale = "bayesian.bridge.scale.0",
      alpha = "sitemodel.shape.0",
      log.kappa = "substmodel.kappa.unres.0",
      rootHeight = "tree.root_height.0"
    )
  )) %>%
  mutate(type = "torchtree") %>%
  gather("param", "value",-type)

other = rbind(beast.other, t.bb.other)
```

```{r}
ggplot(other, aes(value, group = type, colour = type)) +
  geom_density() +
  facet_wrap( ~ param, scales = "free")
```


\newpage

# Normalized rates

```{r}
lookup = setNames(
  c("log.kappa", sapply(1:12, function(x) {
    paste0("glmRandCoefficients", x)
  })),
  c("kappa", "AC", "AG", "AT", "CG", "CT", "GT", "CA", "GA", "TA", "GC", "TC", "TG")
)
df.beast.rates = df.beast %>%
  select(starts_with("glmRandCoefficients") | log.kappa) %>%
  exp() %>%
  rename(all_of(lookup)) %>%
  # mutate(rateSum = select(., AC, AG, AT, CG, CT, GT, CA, GA, TA, GC, TC, TG) %>%
  #     rowSums()) %>%
  # mutate_at(
  #     c("AC", "AG", "AT", "CG", "CT", "GT", "CA", "GA", "TA", "GC", "TC", "TG"),
  #     ~ . / rateSum
  # ) %>%
  mutate(
    AC = AC * freqs[2], AG = AG * freqs[3] * kappa, AT = AT * freqs[4],
    CG = CG * freqs[3], CT = CT * freqs[4] * kappa, GT = GT * freqs[4]
  ) %>%
  mutate(
    CA = CA * freqs[1], GA = GA * freqs[1] * kappa, TA = TA * freqs[1],
    GC = GC * freqs[2], TC = TC * freqs[2] * kappa, TG = TG * freqs[3]
  ) %>%
  mutate(norm = ((AC + AG + AT) * freqs[1] + (CA + CG + CT) * freqs[2] +
    (GA + GC + GT) * freqs[3] + (TA + TC + TG) * freqs[4])) %>%
  mutate_at(
    c("AC", "AG", "AT", "CG", "CT", "GT", "CA", "GA", "TA", "GC", "TC", "TG"),
    ~ . / norm
  ) %>%
  select(-c(kappa, norm)) %>%
  mutate(type = "HKY-RE-beast") %>%
  gather("param", "value", -type)

lookup = sapply(0:11, function(x) {
  paste0("substmodel.kappa.", x)
})
names(lookup) = c("AC", "AG", "AT", "CG", "CT", "GT", "CA", "GA", "TA", "GC", "TC", "TG")

df.t.rates = df.t %>%
  select(matches("substmodel.kappa.\\d+")) %>%
  rename(all_of(lookup)) %>%
  # mutate(rateSum = select(., AC, AG, AT, CG, CT, GT, CA, GA, TA, GC, TC, TG) %>%
  #     rowSums()) %>%
  # mutate_at(
  #     c("AC", "AG", "AT", "CG", "CT", "GT", "CA", "GA", "TA", "GC", "TC", "TG"),
  #     ~ . / rateSum
  # ) %>%
  mutate(
    AC = AC * freqs[2], AG = AG * freqs[3], AT = AT * freqs[4],
    CG = CG * freqs[3], CT = CT * freqs[4], GT = GT * freqs[4]
  ) %>%
  mutate(
    CA = CA * freqs[1], GA = GA * freqs[1], TA = TA * freqs[1],
    GC = GC * freqs[2], TC = TC * freqs[2], TG = TG * freqs[3]
  ) %>%
  mutate(norm = ((AC + AG + AT) * freqs[1] + (CA + CG + CT) * freqs[2] +
    (GA + GC + GT) * freqs[3] + (TA + TC + TG) * freqs[4])) %>%
  mutate_at(
    c("AC", "AG", "AT", "CG", "CT", "GT", "CA", "GA", "TA", "GC", "TC", "TG"),
    ~ . / norm
  ) %>%
  select(-norm) %>%
  mutate(type = "HKY-RE") %>%
  gather("param", "value", -type)

df = rbind(df.t.gtr.rates, df.t.rates, df.beast.rates, df.beast.gtr.rates)
```

```{r}
df$param <- factor(df$param,
  levels = c("AC", "CA", "AG", "GA", "AT", "TA", "CG", "GC", "CT", "TC", "GT", "TG")
)
ggplot(df, aes(value, param, color = type)) +
  geom_boxplot(outlier.shape = NA) +
  coord_flip() +
  scale_x_continuous(trans = "log2", breaks = c(0.01, 0.05, 0.2, 0.5, 2.0)) +
  xlab("Normalized rate") +
  ylab("Substitution type")
```

```{r nice-plot}
temp = data.frame(
  param = c("AC", "CA", "AG", "GA", "AT", "TA", "CG", "GC", "CT", "TC", "GT", "TG"),
  index = c(1, 1, 2, 2, 3, 3, 4, 4, 5, 5, 6, 6)
)
df = df %>% left_join(temp, by = join_by(param))
ggplot(df, aes(value, param, color = type, fill = type)) +
  theme_bw() +
  theme(
    legend.position = "top",
    legend.title = element_blank(),
    strip.background = element_blank(),
    strip.text.x = element_blank()
  ) +
  geom_boxplot(outlier.shape = NA, alpha = 0.5) +
  coord_flip() +
  scale_x_continuous(trans = "log2",
                     breaks = c(0.01, 0.02, 0.05, 0.1, 0.2, 0.5, 1.0, 2.0)) +
  xlab("Normalized rate") +
  ylab("Substitution type") +
  # scale_colour_manual(values = c("#E69F00", "#009E73"), name = "") +
  # scale_fill_manual(values = c("#E69F00", "#009E73"), name = "") +
  facet_grid(. ~ index, drop = TRUE, scale = "free")
```

\newpage

# Bayes factor

Let $\phi = \tau^{-\alpha}$ and $\phi \sim Gamma(1, rate=0.5)$.
Using the change of variable formula the pdf of $\tau$ is

$$p(\tau) = p_\phi(\phi) |d\phi/d\tau| = p_\phi(\tau^{-\alpha}) |-\alpha \tau^{-\alpha-1}|$$

The prior density at $\Delta_{ij} = \epsilon_{ji} - \epsilon_{ij}=0$ given $\tau$ is

$$p(\Delta_{ij}=0|\tau) = \frac{\alpha}{2^{1+1/\alpha} \tau \Gamma(1/\alpha)}$$

Integrating $\tau$ we get

$$p(\Delta_{ij}=0) = \int_0^\infty p(\Delta_{ij}=0|\tau) p(\tau) d\tau.$$

The Bayes Factor in favor of $\Delta_{ij} = 0$ (Model 0, against Model 1 where $\Delta_{ij}$ is a free parameter) is the ratio of the posterior density to the prior density at $\Delta_{ij} = 0$,

$$BF_{01} = \frac{p(\Delta_{ij}=0 | y)}{p(\Delta_{ij}=0)} \approx \frac{q(\Delta_{ij}=0)}{p(\Delta_{ij}=0)},$$

where $q$ is a normal distribution with mean $\mu_{ji}-\mu_{ij}$ and variance $\sigma_{ij}^2 + \sigma_{ji}^2$.
$\mu_{ij}$ and $\sigma_{ij}$ are the mean and standard deviation of the normal variational approximation of $\epsilon_{ij}$.

```{r}
j = fromJSON(file.path(resultsdir, "torchtree", "HKY-RE", "skyglide", "ELBO", "checkpoint.json"))
loc = j$tensor[[match("var.Normal.substmodel.kappa.random.loc", j$id)]]
log_scale = j$tensor[[match("var.Normal.substmodel.kappa.random.scale.unres", j$id)]]
scale = exp(log_scale)

post0 = dnorm(0., loc[7:12] - loc[1:6], sd = sqrt(scale[1:6]**2 + scale[7:12]**2))

alpha = 0.25
prior.tau.0 <- function(tau) {
  alpha / (2. ** (1. + 1. / alpha) * tau * gamma(1. / alpha)) *
    dgamma(tau ** (-alpha), 1.0, rate = 0.5) * alpha * tau ** (-alpha - 1)
}
prior0 = integrate(prior.tau.0, lower = 0, upper = Inf)
df = data.frame(
  Substitution = c("A<->C", "A<->G", "A<->T", "C<->G", "C<->T", "G<->T"),
  BF01 = post0 / prior0$value,
  BF10 = prior0$value/post0,
  logBF10 = log10(prior0$value/post0),
  posterior = post0,
  prior = prior0$value
)
kable(df)
```

```{r}
coeffs = df.beast%>% select(starts_with("glmRandCoefficients"))
locs = coeffs[, glmRandCoefficients1:glmRandCoefficients6] - coeffs[, glmRandCoefficients7:glmRandCoefficients12]
beast.post0 = numeric(ncol(locs))
for(i in 1:ncol(locs)){
  values = locs[[paste0("glmRandCoefficients", i)]]
  #d = density(values)
  #beast.post0[i] = approx(d$x,d$y,xout=0.0)$y
  beast.post0[i] = dnorm(0.0, mean(-values), sd(-values))
}
```

```{r}
temp = locs %>% gather("param", "value")
ggplot(temp, aes(value, group = param)) +
  geom_density() +
  facet_wrap(~param, scales = "free")
```

```{r}
df = data.frame(
  Substitution = c("A<->C", "A<->G", "A<->T", "C<->G", "C<->T", "G<->T"),
  BF01 = beast.post0 / prior0$value,
  BF10 = prior0$value / beast.post0,
  logBF10 = log10(prior0$value / beast.post0),
  posterior = beast.post0,
  prior = prior0$value
)
kable(df)
```

```{r}
df = data.frame(
  Substitution = c("A<->C", "A<->G", "A<->T", "C<->G", "C<->T", "G<->T"),
  BEAST = log10(prior0$value / beast.post0),
  torchtree = log10(prior0$value / post0)
)
kable(df)
```

```{r, eval=FALSE,include=FALSE}
plot(seq(0, 1000000, 1), prior.tau.0(seq(0, 1000000, 1)))
```
